---
title: "Demonstrating Categorical Predictors in Discordant-Kinship Regressions"
author: Yoo Ri Hwang, Jonathan Trattner, and S. Mason Garrison
bibliography: references.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{categorical_predictors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

# Introduction


This vignette provides a comprehensive guide to incorporating categorical predictors in discordant-kinship regression analyses using the 'discord' package. The focus will be on the handling of different types of categorical variables—mixed variables and between-dyad variables—within the discordant-kinship framework.

## Data and Research Question

We will be using a subset of the 1979 National Longitudinal Survey of Youth (NLSY79) to explore whether sex is significant significant predictor for SES at age 40, using different categorical variable coding schemes.


# Data cleaning

First, we load the package and the data.
 
```{r setup, message = FALSE}

# For easy data manipulation
library(dplyr)
# For kinship linkages
library(NlsyLinks)
# For discordant-kinship regression
library(discord)
# pipe
library(magrittr)

#data
data(data_flu_ses)
```

The code chunk below is from Regression demonstration vignette in this package.
For explanation of basic logic for this, please refer to the [regression demonstration vignette](https://github.com/R-Computing-Lab/discord/blob/main/vignettes/regression.Rmd).

```{r set-df-link}


# for reproducibility 
set.seed(2023)

link_vars <- c("S00_H40", "RACE", "SEX","FLU_total")


# Specify NLSY database and kin relatedness 

link_pairs <- Links79PairExpanded %>%
  filter(RelationshipPath == "Gen1Housemates" & RFull == 0.5)

df_link <- CreatePairLinksSingleEntered(outcomeDataset = data_flu_ses,
                                        linksPairDataset = link_pairs,
                                        outcomeNames = link_vars)



# We removed the pair when the Dependent Variable is missing. 

df_link <- df_link %>%
           filter(!is.na(S00_H40_S1) & !is.na(S00_H40_S2))

# We only pick one pair per one household so that we don't violate independence assumption.


df_link<-df_link%>%
  group_by(ExtendedID) %>%
  slice_sample() %>%
  ungroup()

```


# Handling Categorical Predictors



## The general logic of discordant-kinship regressions

If the predictor is a continuous variable, the DV (the difference score of outcome variable) is regressed on the mean score of outcome variable, the mean score of the predictor, and the difference score of the predictor. However, it is impossible to calculate the mean score and difference score of categorical variable. This vignette introduces how to deal with categorical predictors in discordant-kinship regressions. In this example, we use sex and race as categorical variables for the demonstrating purpose. We referred to  @kenny2006  for idea about how to handle categorical predictors in dyadic analysis, and @Hwang2022 for the ways corporate that idea into discordant-kinship regression.




## Mixed Variables: Gender as an Example

In kinship regression, a variable can be at the within dyad level, the between-dyad level, or a combination. This combination is called a "mixed variable", meaning that it varies both within and between dyads. Gender is a prototypical example of a mixed variable <confirm>.

The categorical predictor is a mixed variable when variances exists within- and between-dyads. 
For example, the sex is a mixed variable in that individuals in the pair can be male or female, regardless of their dyad. 


To prepare the data for analysis, we use the `discord_data()` function.


The 'discord_data()' function restructures the paired data. 
In the pair, the one with the higher DV is assigned to be "_1" and the other one is assigned to be "_2".

```{r}


cat_sex <- discord_data(
            data = df_link,
            outcome = "S00_H40",
            sex="SEX",
            race="RACE",
            demographics = "sex",
            predictors=NULL,
            pair_identifiers = c("_S1","_S2"),
            added_coding="added_coding")
```

A random slice of this data looks as following:

```{r sex, echo = FALSE, eval = knitr::is_html_output(),error=FALSE}

cat_sex %>%
slice(1:500) %>%
    slice_sample(n = 6) %>%
    kableExtra::kbl('html', align = "c")%>%
    kableExtra::kable_styling(full_width = FALSE)
```


The sex compositions in the data are as following:

```{r preview-sex, echo = FALSE, eval = knitr::is_html_output()}

cat_sex %>%
  group_by(SEX_1, SEX_2) %>%
  summarize( n(),.groups = 'drop') %>%
  kableExtra::kbl('html', align = "c",col.names=c("SEX_1","SEX_2","sample_size"))%>%
  kableExtra::kable_styling(full_width = FALSE) 

```



Considering the DV (the difference score of outcome variable in the pair) is a between-dyad variable, we can make the sex variable as a between-dyad variable as well. 

This means that rather than using individual's sex as a predictor in the regressions, using sex-composition of the pair in the regressions can be yield more meaningful results. If we put the individuals sex in the pair as a predictor, that is individual level's predictor while DV is not individual level variable. However, by forcing sex variable to be between-dyad variable (by using gender-composition variable), results can be more interpretable and meaningful. 

So, how to force sex variable to be sex-composition variable? The above table shows all the possible combination of sex composition variable: 1) 00, 2) 01, 3) 10, and 4) 11. However, it is hard to believe that "01" and "10" distinction is meaningful unless there are strong theoretical reasons behind, because the "_S1" and "_S2" was assigned by DV value of the participants. Thus "01" pairs means that the one with higher DV in the pair was "0" and the other one was "1", and "10" pairs means that the one with higher DV in the pair was "1" and the other one was "0". 

Thus, we can utilize the following sex-composition variables. First, we can use sex-composition variable that have three factors: 1) "00",2) "10" and "01", and 3) "11". Or, we can also use sex-composition variable that have two factors: 1) "00" and "11", 2) "01" and "10". The former option can be named as "multi-match" and the latter option can be named as "binary-match". 


The 'discord_data()' function automatically generated binary match and multi match for the sex variable. 

The binary-match variable indicates whether the pair is the same-sex pair (coded as "1") or mixed-sex pair (coded as "0").

The multi-match variable indicates whether the sex of the two people in the same pair is "1" (coded as "1") or "0" (coded as "0") or the pair is mixed-sex pair (coded as "mixed"). 

By doing so, the sex variable (which was initially a mixed variable) becomes between-dyad variable as following:


```{r sex-compositions, echo = FALSE, eval = knitr::is_html_output()}

cat_sex %>%
  group_by(SEX_binarymatch,SEX_multimatch, SEX_1, SEX_2) %>%
  summarize( n(),.groups = 'drop') %>%
  kableExtra::kbl('html', align = "c",col.names=c("binary","multi", "SEX_1","SEX_2","sample_size"))%>%
  kableExtra::kable_styling(full_width = FALSE) 
```

Researchers can choose between two options based on their research question. 

For example, if research question focuses on the difference between same-sex pairs and mixed-sex pairs, binary-match variable can be a good option. If research question focuses on the difference between male-male pairs, female-female pairs, and mixed-sex pairs, multi-match can be a good option. 

The regression model with the binary match sex variable can be conducted as such:

```{r echo = FALSE, eval = knitr::is_html_output()}

 lm(S00_H40_diff ~  S00_H40_mean + SEX_binarymatch,
    data = cat_sex)%>%
  broom::tidy() %>%
  mutate(p.value = scales::pvalue(p.value, add_p = TRUE),
         across(.cols = where(is.numeric), ~round(.x, 3))) %>%
  rename("Standard Error" = std.error,
         "T Statistic" = statistic) %>%
  rename_with(~snakecase::to_title_case(.x)) %>%
  kableExtra::kbl('html', align = "c") %>%
  kableExtra::kable_styling() %>%
  kableExtra::column_spec(1:5, extra_css = "text-align: center;")

```

The mean score of SES at 40 was a significant control variable (p <.001). The mean score of SES of the pairs at 40 is negatively associated with the difference score of SES between the pairs, controlling for other variable (in this case, the Sex-composition variable). It is estimated that for one unit increase of the mean score of SES at 40, DV (the difference score of SES at 40) is expected to decrease approximately 0.088. 


The SEX_binary match was not a significant predictor (p = 0.886) when controlling for the mean score of SES at 40. There was not significant differences between same-sex pairs and mixed-sex pairs in the difference score of SES at 40 (DV). This means that the difference between same-sex pairs and mixed-sex pairs were not significantly predicts the difference score of SES at 40 (DV) in the pair when controlling for the mean score of SEX at 40. 

It is estimated that, compared to the reference group (the mixed-sex pairs), the "1" group (the same-sex pairs) would have approximately 0.107 higher difference score of SES at 40 (DV), when controlling for the mean score of SES at 40. However, this coefficient is not significant, so it is not advisable to interpret the coefficient. 


The regression model with the multi match sex variable can be conducted as such:

```{r echo = FALSE, eval = knitr::is_html_output()}

 lm(S00_H40_diff ~ S00_H40_mean + SEX_multimatch,
    data = cat_sex) %>%
  broom::tidy() %>%
  mutate(p.value = scales::pvalue(p.value, add_p = TRUE),
         across(.cols = where(is.numeric), ~round(.x, 3))) %>%
  rename("Standard Error" = std.error,
         "T Statistic" = statistic) %>%
  rename_with(~snakecase::to_title_case(.x)) %>%
  kableExtra::kbl('html', align = "c") %>%
  kableExtra::kable_styling() %>%
  kableExtra::column_spec(1:5, extra_css = "text-align: center;")

```

The mean score of SES at 40 was a significant control variable (p <.001). The mean score of SES of the pairs at 40 is negatively associated with the difference score of SES between the pairs, controlling for other variable (in this case, the Sex-composition variables). It is estimated that for one unit increase of the mean score of SES at 40, DV (the difference score of SES at 40) is expected to decrease approximately 0.087. 


There was no significant difference between "00" pairs (the pairs that two people's sex are both 0 ) and "11" pairs (the pairs that two people's sex are both 1) (p = 0.705) to predict the DV. Similarly, there was no significant differences between mixed-sex pairs and "00" pairs (p = 0.916).

The coefficient 0.399 is the difference between the expected DV (the difference score of SES at 40) for the reference group (in this case, the "00" pairs) and the "11" pairs. 
The coefficient 0.097 is the difference between the expected DV (the difference score of SES at 40) for the reference group (in this case, the "00" pairs) and the mixed-sex pairs. However, these coefficients were not significant, so it is not advisable to interpret the coefficients. 



## Between-dyad variable: Race as an Example

The categorical predictor is a between-dyad variable when the two people in the same dyad share the same score (or same response). For example, the length of the marriage is a between-dyad variable if the study consists of married-couple samples. Race can be a between-dyad variable if the study consists of same-race sibling pairs.  

For demonstration purpose, we force the race variable to be between-dyad variable, meaning that we only include the same-race pairs in the data. 

By using 'discord_data()' function, we can prepare data for the analysis as following:

```{r}

# for reproducibility 

set.seed(2023)

cat_race <- discord_data(
            data = df_link,
            outcome = "S00_H40",
            predictors = NULL,
            sex="SEX",
            race="RACE",
            demographics = "race",
            pair_identifiers = c("_S1","_S2"),
            added_coding="added_coding") %>%
# forcing race to be a between-dyad variable for the demonstration purpose
            filter(RACE_1==RACE_2)

```

The race compositions in the data set are:

```{r preview-race, echo = FALSE, eval = knitr::is_html_output()}
cat_race %>%
  group_by(RACE_binarymatch, RACE_multimatch, RACE_1, RACE_2)%>%
  summarize( n(),.groups = 'drop')%>%
  kableExtra::kbl('html', align = "c",col.names=c("RACE_binarymatch", "RACE_multimatch", "RACE_1","RACE_2","sample_size"))%>%
  kableExtra::kable_styling(full_width = FALSE) 
  
```

The binary-match variable indicates whether the pair is the same-race pair (coded as "1") or mixed-race pair (coded as "0"). Because the race variable is a between-dyad variable, all the pairs have a value of "1" in this example. 

The multi-match variable indicates whether the race of the two people in the same pair is both "1" (coded as "1") or "0" (coded as "0") or mixed (coded as "mixed"). No pair have "mixed" value because race is a between-dyad variable in this example. 

The regression model with multi-match race variable as a  predictor can be conducted  as such:

```{r echo = FALSE, eval = knitr::is_html_output()}

 lm(S00_H40_diff ~  S00_H40_mean + RACE_multimatch,
    data = cat_race) %>%
  broom::tidy() %>%
  mutate(p.value = scales::pvalue(p.value, add_p = TRUE),
         across(.cols = where(is.numeric), ~round(.x, 3))) %>%
  rename("Standard Error" = std.error,
         "T Statistic" = statistic) %>%
  rename_with(~snakecase::to_title_case(.x)) %>%
  kableExtra::kbl('html', align = "c") %>%
  kableExtra::kable_styling() %>%
  kableExtra::column_spec(1:5, extra_css = "text-align: center;")

```

The mean score of SES at 40 was significant control variable (p < .001). The mean score of SES of the pairs at 40 is negatively associated with the difference score of SES between the pairs, controlling for other variable (in this case, the Sex-composition variable). It is estimated that for one unit increase of the mean score of SES at 40, DV (the difference score of SES at 40) is expected to decrease approximately 0.074.


The RACE_multimatch variable was significant predictor of DV (the difference score of SES at 40; p = .004) after controlling for the mean score of SES at 40. This means that the difference between the "00" pairs and "11" pairs significantly predicts the DV. Specifically, compared to the reference group (the "00"pairs), "11" pairs are expected to have have approximately 2.22 higher difference score of SES at 40 (DV).  

## When we have two categorical predictors

We can include both sex and race as predictors as well. 

First, we restructure the data for the kinship-discordant regression using the 'discord_data()'function.

```{r}



cat_both <- discord_data(
            data = df_link,
            outcome = "S00_H40",
            predictors = NULL ,
            sex="SEX",
            race="RACE",
            demographics = "both",
            pair_identifiers = c("_S1","_S2"),
            added_coding="added_coding") %>%
# forcing race to be between-dyad variable for demonstration purpose
            filter(RACE_1==RACE_2)

```


 We can perform a regression using the binary-match sex variable and multi-match race variable as such:
 
```{r echo = FALSE, eval = knitr::is_html_output()}
 
 lm(S00_H40_diff ~  S00_H40_mean + RACE_multimatch + SEX_binarymatch,
    data = cat_both) %>%
  broom::tidy() %>%
  mutate(p.value = scales::pvalue(p.value, add_p = TRUE),
         across(.cols = where(is.numeric), ~round(.x, 3))) %>%
  rename("Standard Error" = std.error,
         "T Statistic" = statistic) %>%
  rename_with(~snakecase::to_title_case(.x)) %>%
  kableExtra::kbl('html', align = "c") %>%
  kableExtra::kable_styling() %>%
  kableExtra::column_spec(1:5, extra_css = "text-align: center;")

```


The mean score of SES at 40 was a significant control variable (p <.001). The mean score of SES of the pairs at 40 is negatively associated with the difference score of SES between the pairs, controlling for other variable (in this case, the Sex-composition and race-composition variable). It is estimated that for one unit increase of the mean score of SES at 40, DV (the difference score of SES at 40) is expected to decrease approximately 0.074. 

The SEX_binary match was not a significant predictor (p = 0.938) when controlling for other variable (i.e., the mean score of SES at 40 and the race-composition variable). This means that the difference between same-sex pairs and mixed-sex pairs were not significantly predicts the difference score of SES at 40 (DV) in the pair when controlling for the mean score of SEX at 40. Compared to the reference group (mixed-sex pairs), it is estimated that the the same-sex pairs to have approximately 0.058 higher difference score of SES at 40 (DV), when controlling for the mean score of SES at 40 and race. However, this variable was not significant, so it is not advisable to interpret the coefficient. 

The RACE_multi match was significant predictor (p = .004). This means that there is a significant difference between "00" race pairs and "11" race pairs in the DV when controlling for the model covariates (i.e., Sex composition variable and the mean score of SES at 40). Specifically, compared to the reference group ("00" race pair), "11" race pairs were expected to have approximately 2.22 higher difference score of SES at 40 (DV). 




We can perform a regression analysis using multi-match sex and race variables as well:

```{r echo = FALSE, eval = knitr::is_html_output()}


 lm(S00_H40_diff ~  S00_H40_mean + RACE_multimatch + SEX_multimatch,
    data = cat_both) %>%
  broom::tidy() %>%
  mutate(p.value = scales::pvalue(p.value, add_p = TRUE),
         across(.cols = where(is.numeric), ~round(.x, 3))) %>%
  rename("Standard Error" = std.error,
         "T Statistic" = statistic) %>%
  rename_with(~snakecase::to_title_case(.x)) %>%
  kableExtra::kbl('html', align = "c") %>%
  kableExtra::kable_styling() %>%
  kableExtra::column_spec(1:5, extra_css = "text-align: center;")

```


The mean score of SES at 40 was a significant control variable (p <.001). The mean score of SES of the pairs at 40 is negatively associated with the difference score of SES between the pairs, controlling for other variable (in this case, the Sex-composition and race-composition variable). It is estimated that for one unit increase of the mean score of SES at 40, DV (the difference score of SES at 40) is expected to decrease approximately 0.074. 

The SEX-composition variables were not significant predictors (ps > .05) when controlling for other variable (i.e., the mean score of SES at 40 and the race-composition variable). The coefficient 0.432 is the difference between the expected DV (the difference score of SES at 40) for the reference group (in this case, the “00” pairs) and the “11” pairs. The coefficient 0.163 is the difference between the expected DV (the difference score of SES at 40) for the reference group (in this case, the “00” pairs) and the mixed-sex pairs. However, these coefficients were not significant, so it is not advisable to interpret the coefficients.
 
The RACE_multi match was a significant predictor (p = .004) when controlling for other variables (i.e., the mean score of SES at 40 and the sex-composition variables). This means that there is a significant difference between "00" race pairs and "11" race pairs in the difference score of SES at 40 (DV) when controlling for the model covariates (i.e., Sex composition variables and the mean score of SES at 40). Specifically, compared to the reference group ("00" race pair), the "11" race pairs were expected to have approximately 2.226 higher difference score of SES at 40 (DV). 


# Conclusion

In this vignette, we demonstrated how to include categorical predictors in discordant-kinship regressions using the 'discord' package. 

# References


