---
title: "Kinship Simulations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Kinship Simulations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

This vignette describes kinship simulations with `discord`. The function that takes care of this is `kinsim`. It provides a highly customizable framework for creating biometrically informed multivariate data given the ACE parameters, where *A* is the additive genetic effects, *C* is the common (shared) environmental effects, and *E* is the non-shared (individual-specific) environmental effects.

Below are some use cases for different simulations. Note that the package must be installed and loaded as follows:

```{r setup, eval = FALSE}
# Install/update discord with the release version from CRAN.
install.packages('discord')

# Install/update discord with the development version from GitHub. 
# If devtools is not installed, uncomment the line below.
# install.packages('devtools')
devtools::install_github('R-Computing-Lab/discord')

# Load package
library(discord)
```
