---
title: "Visualizing Sibling Differences"
author: S. Mason Garrison
bibliography: references.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visualizing Sibling Differences}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

# Overview


This vignette recreates the style of a figure in [Garrison and Rodgers (2016)](https://www.sciencedirect.com/science/article/pii/S0160289616300162) using `ggplot2` and synthetic data structured with `discord_data()`. The figure illustrates the patterns of relationships between sibling differences in socioeconomic status (Behavior) and sibling differences in the depressiveness. 


# Data Generation and Preparation

## Data Cleaning



The underlying dataset is the NLSY79, which includes measures of depressiveness vaccination and socioeconomic status (Behavior) for kinship pairs.
As in the regression vignette, we restrict the sample to individuals who are housemates and have a relatedness of 0.5.

For full details on the data processing and kinship link construction, see the [regression vignette](https://r-computing-lab.github.io/discord/articles/regression.html).

<details>
  <summary>Click to expand data preparation code</summary>
```{r, warning=FALSE, message=FALSE}
# Setup: Use discord_data
# Visualizing the Results

library(discord)
library(NlsyLinks)
library(tidyverse)
library(ggplot2)
library(grid)
library(gridExtra)
library(ggExtra)
library(janitor)

data(mz_signif)
data(mz_nonsignif)
set.seed(18)



discord_mz_sig <- discord_data(mz_signif,
    outcome = "y1",
    predictors = "y2",
    id = "id",
    sex = NULL,
    race = NULL,
    demographics = "none",
    pair_identifiers = c("_1", "_2"),
    fast = TRUE
  )  %>%
  rename_with(~ str_replace_all(., c("y1" = "depressiveness", "y2" = "Behavior")), everything()) 

discord_mz_nsig <- discord_data(mz_nonsignif,
    outcome = "y1",
    predictors = "y2",
    id = "id",
    sex = NULL,
    race = NULL,
        demographics = "none",
    pair_identifiers = c("_1", "_2"),
    fast = TRUE
  )   %>%
  rename_with(~ str_replace_all(., c("y1" = "depressiveness", "y2" = "Behavior")), everything()) 

df_mz_signif <- mz_signif  %>%
  rename_with(~ str_replace_all(., c("y1" = "depressiveness", "y2" = "Behavior")), everything()) 

df_mz_nonsignif <- mz_nonsignif  %>%
  rename_with(~ str_replace_all(., c("y1" = "depressiveness", "y2" = "Behavior")), everything()) 

df_used <- df_mz_signif
```
</details>

## Creating the Discord Data

With the data prepared, we restructure it using `discord_data()`.

```{r, warning=FALSE}
library(tidyverse)
library(ggplot2)

df_discord_depressiveness <- discord_data(df_mz_signif,
    outcome = "depressiveness",
    predictors = "Behavior",
    id = "id",
    sex = NULL,
    race = NULL,
    demographics = "none",
    pair_identifiers = c("_1", "_2"),
    fast = TRUE
  )
```

Because we are interested in differences between kin, we create a new variable, `mom_diff_group`, that classifies Behavior differences into three categories: "More Advantaged", "Equally Advantaged", and "Less Advantaged". This variable is later used to group observations in the marginal density plots. They serve to help visualize how the distributions of mean Behavior and mean depressiveness vaccinations differ across these Behavior difference categories.

```{r}
df_discord_depressiveness <- df_discord_depressiveness %>%
  mutate(
    ses_mean_group = factor(
      case_when(
        as.numeric(scale(Behavior_mean)) > 0.5 ~ "More Advantaged",
        as.numeric(scale(Behavior_mean)) < -0.5 ~ "Less Advantaged",
        abs(as.numeric(scale(Behavior_mean))) <= 0.5 ~ "Equally Advantaged"
      ),
      levels = c(
        "Less Advantaged",
        "Equally Advantaged",
        "More Advantaged"
      )
    ),
    # # Classify Difference Grouping
    mom_diff_group = factor(
      case_when(
        as.numeric(scale(Behavior_diff)) > 0.5 ~ "More Advantaged",
        as.numeric(scale(Behavior_diff)) < -0.5 ~ "Less Advantaged",
        abs(as.numeric(scale(Behavior_diff))) <= 0.5 ~ "Equally Advantaged"
      ),
      levels = c(
        "Less Advantaged",
        "Equally Advantaged",
        "More Advantaged"
      )
    )
  )
```




## Setting Up Color Palette

To enhance the visualizations, we define a color palette that reflects the Behavior differences between siblings. Here, we use a gradient that transitions from red to blue. This color scheme helps to intuitively convey the direction and magnitude of Behavior differences, with red indicating lower Behavior and blue indicating higher Behavior. Missing values are represented in purple, which is approximately midway between red and blue.


```{r}
# Create a color palette for the shading
color_shading_4 <- c("firebrick4", "firebrick1", "dodgerblue1", "dodgerblue4")
color_na <- "#AD78B6" # purple for missing values

color_shading_3 <- c(color_shading_4[2], color_na, color_shading_4[3])

# Determine the range of Behavior differences for color scaling
max_val_scale <- max(abs(df_discord_depressiveness$Behavior_diff), na.rm = TRUE)

# values <- seq(-max_val_scale, max_val_scale, length = length(color_shading_4))
```


# Plotting the Results

## Individual Level Plot

This plot is for looking at individual-level data rather than sibling pair means or differences. It provides context for understanding the relationship between Behavior and depressiveness vaccinations before examining sibling differences.

This scatter plot shows individual Behavior at age 40 against individual depressiveness vaccinations. Point color indicates the Behavior difference between siblings. 

The first step is to create the base plot with sibling 1 data. In the next code block, we add sibling 2 data to the plot. 

```{r individual, echo=TRUE, message=FALSE}
# Individual level plot
plot_indiv <- plot_indiv_sib1 <- ggplot(
  df_used,
  aes(
    x = Behavior_1,
    y = depressiveness_1,
    color = Behavior_1 - Behavior_2
  )
) +
  geom_point(
    size = 0.8, alpha = 0.8, na.rm = TRUE,
    position = position_jitter(width = 0.2, height = 0.2)
  ) +
  geom_smooth(method = "lm", se = FALSE, color = "black")
```

But if we only plot sibling 1, we can visualize that first.

```{r echo=TRUE, message=FALSE, warning=FALSE}
plot_indiv_sib1 +
  scale_colour_gradientn(
    name = "Kin\nDifferences\nin Behavior",
    colours = color_shading_4,
    na.value = color_na,
    values = scales::rescale(c(-max_val_scale, max_val_scale))
  ) +
  labs(
    x = "Behavior",
    y = "Depressiveness"
  ) +
  theme_minimal() +
  ggtitle("Individual Level Plot: Kin 1 Only") +
  theme(plot.title = element_text(hjust = 0.5))
```

Now, we add sibling 2 to the plot, using the same color scheme to indicate Behavior differences between siblings. 

```{r echo=TRUE, message=FALSE, warning=FALSE}
plot_indiv <- plot_indiv_sib1 +
  # added sibling 2 to the plot
  geom_point(
    size = 0.8, alpha = 0.8, na.rm = TRUE,
    position = position_jitter(width = 0.2, height = 0.2),
    aes(
      x = Behavior_2,
      y = depressiveness_2,
      color = Behavior_2 - Behavior_1 # this reverses the color difference so sibling 2 points use the opposite color gradient compared to sibling 1, making it visually clear which sibling is being represented and how their Behavior difference is encoded
    )
  ) +
  scale_colour_gradientn(
    name = "Kin \nDifferences\nin Behavior",
    colours = color_shading_4,
    na.value = color_na,
    values = scales::rescale(c(-max_val_scale, max_val_scale))
  ) +
  labs(
    x = "Maternal Behavior",
    y = "Child depressiveness"
  ) +
  theme_minimal() #+
#  theme(legend.position = "none")

individual_level_plot_b <- plot_indiv +
  ggtitle("Individual Level Panel B") +
  theme(plot.title = element_text(hjust = 0.5)) 


  ggsave("vignettes/individual_level_plot_b.png", plot = individual_level_plot_b, width = 7, height = 5, dpi = 300)
```


The individual-level plot shows a positive association between Behavior and depressiveness vaccinations. Higher Behavior individuals tend to have higher depressiveness vaccination rates. The color gradient indicates the Behavior difference between siblings, providing additional context for interpreting the data.

```{r}
plot_indiv_s00 <- ggplot(
  df_mz_nonsignif,
  aes(
    x = Behavior_1,
    y = Behavior_2,
    color = Behavior_1 - Behavior_2
  )
) +
  geom_point(
    size = 0.8, alpha = 0.8, na.rm = TRUE,
    position = position_jitter(width = 0.2, height = 0.2)
  ) +
  geom_smooth(method = "lm", se = FALSE, color = "black")


plot_indiv_s00 +
  scale_colour_gradientn(
    name = "Sibling\nDifferences\nin Behavior",
    colours = color_shading_4,
    na.value = color_na,
    values = scales::rescale(c(-max_val_scale, max_val_scale))
  ) +
  labs(
    x = "Maternal Behavior (Kin 1)",
    y = "Maternal Behavior (Kin 2)"
  ) +
  theme_minimal() +
  ggtitle("Individual Level Plot: Behavior Comparison Between Siblings") +
  theme(plot.title = element_text(hjust = 0.5))
```


```{r}
plot_indiv_depressiveness <- ggplot(
  df_mz_nonsignif,
  aes(
    x = depressiveness_1,
    y = depressiveness_2,
    color = Behavior_1 - Behavior_2
  )
) +
  geom_point(
    size = 0.8, alpha = 0.8, na.rm = TRUE,
    position = position_jitter(width = 0.2, height = 0.2)
  ) +
  geom_smooth(method = "lm", se = FALSE, color = "black")


plot_indiv_depressiveness +
  scale_colour_gradientn(
    name = "Differences\nin Depressiveness",
    colours = color_shading_4,
    na.value = color_na,
    values = scales::rescale(c(-max_val_scale, max_val_scale))
  ) +
  labs(
    x = "Depressiveness (Kin 1)",
    y = "Depressiveness (Kin 2)"
  ) +
  theme_minimal() +
  ggtitle("Individual Level Plot: Depressiveness Comparison Between Kin") +
  theme(plot.title = element_text(hjust = 0.5))
```


## Family Level Plots

### Between Family Plots

This section creates a between-family plot that visualizes mean Behavior at age 40 against mean depressiveness vaccinations for sibling pairs. Points are colored based on the Behavior difference between siblings. Each point represents a sibling pair, with the x-axis showing the pair's average Behavior and the y-axis showing the average number of depressiveness vaccinations.


```{r scatter, message=FALSE, include=FALSE, echo=TRUE}
# Main scatter plot
plot_btwn <- ggplot(df_discord_depressiveness, aes(
  x = Behavior_mean,
  y = depressiveness_mean,
  color = mom_diff_group,
)) +
  geom_point( # this layer creates invisible points to all the marginal plots to align correctly
    size = 1.8, alpha = 0.0, na.rm = TRUE,
    shape = 21,
    position = position_jitter(width = 0.2, height = 0.2, seed = 1234)
  ) +
  geom_point(
    size = 1.8, alpha = 0.8, na.rm = TRUE,
    shape = 21,
    aes(
      fill = Behavior_diff,
      colour = mom_diff_group
    ),
    group = 1,
    position = position_jitter(width = 0.2, height = 0.2, seed = 1234)
  ) +
  geom_smooth(
    method = "lm", se = FALSE, color = "black",
    fill = "black",
    group = 1
  ) +
  scale_fill_gradientn(
    name = "Sibling\nDifferences\nin Behavior",
    colours = color_shading_4,
    na.value = color_na,
    values = scales::rescale(c(
      -max_val_scale,
      0,
      max_val_scale
    ))
  ) +
  scale_color_manual(
    values = color_shading_3,
  ) +
  theme_minimal() +
  theme(legend.position = "left") +
  labs(x = "Mean Behavior", y = "Mean depressiveness")
```

```{r echo=TRUE, message=FALSE}
plot_btwn
```

As you can see, the base scatter plot shows a positive association between mean Behavior and mean depressiveness vaccinations. Higher average Behavior among sibling pairs is associated with higher average depressiveness vaccination rates. Marginal plots can be added to further illustrate the distributions of mean Behavior and mean depressiveness vaccinations. Below is the code to add marginal plots using the `ggExtra` package. Note that the marginal plots are histograms, density plots, or boxplots that show the distribution of mean Behavior and mean depressiveness vaccinations, grouped by the Behavior difference category.


```{r echo=TRUE, message=FALSE}
# Set relative size of marginal plots (main plot 10x bigger than marginals)
ggMarginal(plot_btwn,
  type = "histogram", size = 10, # groupColour = TRUE,
  groupFill = T,
  fill = color_shading_3
)
```

```{r echo=TRUE, message=FALSE}
ggMarginal(plot_btwn, type = "density", size = 10, groupColour = F, groupFill = T, aes(
  color = mom_diff_group,
  fill = mom_diff_group,
  alpha = 0.95
))
```

```{r echo=TRUE, message=FALSE}
ggMarginal(plot_btwn, type = "boxplot", size = 10, groupColour = F, groupFill = T)
```

Like the individual-level plot, these between-family plot shows a positive association between mean Behavior and mean depressiveness vaccinations. Higher average Behavior among sibling pairs is associated with higher average depressiveness vaccination rates. The marginal plots further illustrate the distributions of mean Behavior and mean depressiveness vaccinations, grouped by the Behavior difference category. This helps to visualize how the distributions differ across sibling pairs with varying Behavior differences.


### Adding Marginal Density Plots

An alternative approach is to create marginal density plots separately and arrange them alongside the main scatter plot using `gridExtra::grid.arrange()`.

```{r plot-raw-data, message=FALSE}
# Marginal X density (Behavior mean)
plot_xdensity <- ggplot(df_discord_depressiveness, aes(
  x = Behavior_mean,
  group = mom_diff_group,
  color = mom_diff_group
)) +
  geom_density(adjust = 2, linewidth = 1, fill = NA) +
  scale_colour_manual(
    name = "Sibling\nDifferences\nin Behavior",
    values = color_shading_3
  ) +
  theme_minimal() +
  theme(
    legend.position = "left",
    axis.title.y = element_blank(),
    axis.text.y = element_blank(),
    legend.title = element_text(size = 8),
    legend.text = element_text(size = 6)
  ) +
  labs(x = NULL, y = NULL)
```

And for the Y density plot:

```{r}
# Marginal Y density (depressiveness mean)
plot_ydensity <- ggplot(df_discord_depressiveness, aes(
  x = depressiveness_mean,
  group = mom_diff_group,
  color = mom_diff_group
)) +
  geom_density(
    adjust = 2,
    linewidth = 1,
    fill = NA
  ) +
  scale_colour_manual(
    values = color_shading_3
  ) +
  coord_flip() +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.title.x = element_blank(),
    axis.text.x = element_blank()
  ) +
  labs(x = NULL, y = NULL)
```

To finalize the marginal plots, we add titles and adjust themes for better presentation.

```{r echo=FALSE, message=FALSE}
# add titles
plot_xdensity + theme_bw() + theme(
  plot.title = element_text(hjust = 0.5, size = 12),
  axis.title.x = element_text(size = 12)
) + labs(title = "Mean Behavior at Age 40")

plot_ydensity + theme_bw() + theme(
  plot.title = element_text(hjust = 0.5, size = 12),
  axis.title.x = element_text(size = 12)
) + labs(title = "Mean depressiveness Vaccinations (2006–2016)")
```

#### Assembling the Final Plot

Finally, we arrange the main scatter plot and the marginal density plots into a cohesive layout using `gridExtra::grid.arrange()`. The x-density plot is placed above the main scatter plot, and the y-density plot is placed to the right of the main scatter plot. We can do this by creating a blank placeholder plot to fill the layout's empty space.

```{r}
# Blank placeholder plot
plot_blank <- ggplot() +
  theme_void()

# Final layout
grid.arrange(
  arrangeGrob(plot_xdensity,
    plot_blank,
    ncol = 2,
    widths = c(4, 1)
  ),
  arrangeGrob(plot_btwn,
    plot_ydensity,
    ncol = 2,
    widths = c(4, 1)
  ),
  heights = c(1.5, 4),
  top = textGrob("Sibling Means in Behavior and depressiveness Vaccinations",
    gp = gpar(
      fontsize = 20,
      font = 3
    )
  )
)
```



### Bonus: Faceting by Behavior Difference Group

Other ways to plot this data include using `facet_wrap` to create separate panels for each Behavior difference group. This can help visualize the differences in depressiveness vaccinations across different Behavior categories.

```{r echo=FALSE, message=FALSE}
plot_btwn + facet_wrap(~mom_diff_group,
  ncol = 1
) +
  theme(legend.position = "none") +
  labs(title = "Between Family Differences in Behavior and depressiveness Vaccinations by Behavior Difference Group")
```

## Within Family Plots

This plot compares differences in Behavior at age 40 to differences in depressiveness vaccinations, with points colored by Behavior difference. Marginal plots are omitted for simplicity but can be added using the same structure.

```{r plot-within-family, echo=TRUE, message=FALSE}
# Within Family
#

# Main scatter plot

# Setup: Use discord_data

# Main scatter plot
plot_within <- ggplot(df_discord_depressiveness, aes(
  x = Behavior_diff,
  y = depressiveness_diff,
  color = mom_diff_group
)) +
  geom_point( # this layer creates invisible points to all the marginal plots to align correctly
    size = 1.8, alpha = 0.0, na.rm = TRUE,
    shape = 21,
    position = position_jitter(width = 0.2, height = 0.2, seed = 1234)
  ) +
  geom_point(
    size = 1.8, alpha = 0.9, na.rm = TRUE,
    shape = 21,
    aes(
      fill = Behavior_diff,
      colour = mom_diff_group
    ),
    group = 1,
    position = position_jitter(width = 0.2, height = 0.2, seed = 1234)
  ) +
  geom_smooth(
    method = "lm",
    se = FALSE,
    color = "black"
  ) +
  scale_fill_gradientn(
    name = "Sibling\nDifferences\nin Behavior",
    colours = color_shading_4,
    na.value = color_na,
    values = scales::rescale(c(
      -max_val_scale,
      0,
      max_val_scale
    ))
  ) +
  scale_color_manual(
    values = color_shading_3,
  ) +
  theme_minimal() +
  theme(legend.position = "left") +
  labs(
    x = "Diff Behavior",
    y = "Diff Outcome"
  )

plot_within
```

Adding marginal plots to the within-family scatter plot can be done similarly to the between-family plot. Below is the code to add marginal plots using the `ggExtra` package.

```{r echo=TRUE, message=FALSE}
# Set relative size of marginal plots (main plot 10x bigger than marginals)
library(ggExtra)
ggMarginal(plot_within,
  type = "histogram", size = 10, # groupColour = TRUE,
  groupFill = T # ,
  #  color = color_shading_3
)
```

```{r echo=TRUE, message=FALSE}
ggMarginal(plot_within, type = "density", size = 10, groupColour = F, groupFill = T, aes(
  color = mom_diff_group,
  fill = mom_diff_group,
  alpha = 0.95
))
```
```{r echo=TRUE, message=FALSE}
ggMarginal(plot_within, type = "boxplot", 
           size = 10, groupColour = F, groupFill = T)
```



You can further facet this plot by the difference in Behavior between kin to see how the relationship varies across different groups. The following code does this and adds a title to the plot.

```{r echo=TRUE, message=FALSE}
plot_within + facet_wrap(~mom_diff_group,
  ncol = 1
) +
  theme(legend.position = "bottom") +
  labs(title = "Within Family Differences in Behavior and Depressiveness")
```

# Conclusion

This vignette demonstrated how to visualize sibling differences in Behavior and Depressiveness vaccinations using discord-structured data. Scatter and density plots highlight associations by Behavior difference group. The use of color to indicate the difference in Behavior between kin adds an additional layer of insight into the data.
