---
title: "Plots"
author: S. Mason Garrison
bibliography: references.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---




This demonstrates how to plot data from the `discord` package. The data used in this example is from the NLSY79 dataset, which contains information on flu vaccination and socioeconomic status (SES) for kinship pairs. These plots are inspired by the work of Garrison and Rodgers (2016), but updated to reflect improvements in the ggplot2 package.

# Data Prep

The data used in this example is from the NLSY79 dataset, which contains information on flu vaccination and socioeconomic status (SES) for kinship pairs. The data is filtered to include only those individuals who are housemates and have a relatedness of 0.5. More details on the kinship link building process can be found in the regression vignette.

## Data Cleaning

```{r, warning=FALSE, message=FALSE}
## Setup: Use discord_data
# Visualizing the Results
library(discord)
library(NlsyLinks)

library(tidyverse)
library(ggplot2)
library(grid)
library(gridExtra)
library(janitor)

# Load the data
data(data_flu_ses)

# Get kinship links for individuals with the following variables:
link_vars <- c(
  "FLU_total", "FLU_2008", "FLU_2010",
  "FLU_2012", "FLU_2014", "FLU_2016",
  "S00_H40", "RACE", "SEX"
)

link_pairs <- Links79PairExpanded %>%
  filter(RelationshipPath == "Gen1Housemates" & RFull == 0.5)


df_link <- CreatePairLinksSingleEntered(
  outcomeDataset = data_flu_ses,
  linksPairDataset = link_pairs,
  outcomeNames = link_vars
)

consistent_kin <- df_link %>%
  group_by(SubjectTag_S1, SubjectTag_S2) %>%
  count(
    FLU_2008_S1, FLU_2010_S1,
    FLU_2012_S1, FLU_2014_S1,
    FLU_2016_S1, FLU_2008_S2,
    FLU_2010_S2, FLU_2012_S2,
    FLU_2014_S2, FLU_2016_S2
  ) %>%
  na.omit()

# Create the flu_modeling_data object with only consistent responders.
# Clean the column names with the {janitor} package.
flu_modeling_data <- semi_join(df_link,
  consistent_kin,
  by = c(
    "SubjectTag_S1",
    "SubjectTag_S2"
  )
) %>%
  clean_names()
```

# Creating the Discord Data

Now that we have prepared the data, we can restructure data with  `discord_data()`. 

```{r}
discord_flu <- discord_data(
  data = flu_modeling_data,
  outcome = "flu_total",
  predictors = "s00_h40",
  id = "extended_id",
  sex = "sex",
  race = "race",
  pair_identifiers = c("_s1", "_s2"),
  demographics = "both"
) %>%
  filter(!is.na(s00_h40_mean), !is.na(flu_total_mean))
```

Because we are interested in the differences between kin, we will create a new variable,  `ses_diff_group`, which classifies the difference in SES into three groups: "More Advantage", "Equally Advantage", and "Less Advantage". We will use this variable to add marginal density plots to the main scatter plot.

```{r}
discord_flu <- discord_flu %>%
  mutate(
    # # Classify Difference Grouping
    ses_diff_group = factor(case_when(s00_h40_diff > 0.33 ~ "More Advantage",
                                      s00_h40_diff < -0.33 ~ "Less Advantage",
                                      abs(s00_h40_diff) <= 0.33 ~ "Equally Advantage"),
                            levels = c("Less Advantage",
                                       "Equally Advantage",
                                       "More Advantage"))
  )

# Create a color palette for the shading
shading <- c("firebrick4", "firebrick1", "dodgerblue1", "dodgerblue4")
max_val <- max(abs(discord_flu$s00_h40_diff), na.rm = TRUE)
values <- seq(-max_val, max_val, length = length(shading))

```

# Plotting the Data

The following code creates a scatter plot of the mean SES at age 40 against the mean flu vaccinations, with points colored by the difference in SES between kin. The marginal density plots show the distribution for each group.

## Core Scatter Plot

```{r scatter, message=FALSE, include=FALSE, echo=TRUE}
# Main scatter plot
main_plot <- ggplot(discord_flu, aes(x = s00_h40_mean,
                                     y = flu_total_mean, colour = s00_h40_diff)) +
  geom_point(size = 0.8, alpha = 0.8, na.rm = TRUE,
             position = position_jitter(w = 0.2, h = 0.2)) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  scale_colour_gradientn(
    name = "Sibling\nDifferences\nin SES",
    colours = shading,
    na.value = "#AD78B6",
    values = scales::rescale(c(min(discord_flu$flu_total_diff, na.rm = TRUE),
                               mean(discord_flu$flu_total_diff, na.rm = TRUE),
                               max(discord_flu$flu_total_diff, na.rm = TRUE)))
  ) +
  theme_minimal() +
  theme( legend.position = "left")+
  labs(x = "Mean SES at Age 40", y = "Mean Flu Vaccinations (2006–2016)")

```

```{r echo=FALSE, message=FALSE}
main_plot
```


# Adding Marginal Density Plots
```{r plot-raw-data, message=FALSE}


# Marginal X density (SES mean)
xdensity <- ggplot(discord_flu, aes(x = s00_h40_mean, group = ses_diff_group, color = ses_diff_group)) +
  geom_density(adjust = 2, linewidth = 1, fill = NA) +
  scale_colour_manual(
    name = "Sibling\nDifferences\nin SES",
    values = c("firebrick1", "#AD78B6", "dodgerblue1")
  ) +
  theme_minimal() +
  theme(legend.position = "left",
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        legend.title = element_text(size = 8),
        legend.text = element_text(size = 6)
        ) +
  labs(x = NULL, y = NULL)

# Marginal Y density (Flu mean)
ydensity <- ggplot(discord_flu, aes(x = flu_total_mean, group = ses_diff_group, color = ses_diff_group)) +
  geom_density(adjust = 2, linewidth = 1, fill = NA) +
  scale_colour_manual(
    values = c("firebrick1", "#AD78B6", "dodgerblue1")
  ) +
  coord_flip() +
  theme_minimal() +
  theme(legend.position = "none",
        axis.title.x = element_blank(),
        axis.text.x = element_blank()) +
  labs(x = NULL, y = NULL)
```
These marginal density plots show the distribution of the mean SES at age 40 and the mean flu vaccinations, respectively. The colors correspond to the difference in SES between kin. Here are the marginal density plots with titles added.


```{r echo=FALSE, message=FALSE}
# add titles
xdensity + theme_bw() + theme(
  plot.title = element_text(hjust = 0.5, size = 12),
  axis.title.x = element_text(size = 12)
) + labs(title = "Mean SES at Age 40")

ydensity + theme_bw() + theme(
  plot.title = element_text(hjust = 0.5, size = 12),
  axis.title.x = element_text(size = 12)
) + labs(title = "Mean Flu Vaccinations (2006–2016)")

```


```{r}
# Blank placeholder plot
blankPlot <- ggplot() +
  theme_void()


    
# Final layout (true layout you originally had: AB above CC)
grid.arrange(
  arrangeGrob(xdensity, blankPlot, ncol = 2, widths = c(4, 1)),
  arrangeGrob(main_plot, ydensity, ncol = 2, widths = c(4, 1)),
  heights = c(1.5, 4),
  top = textGrob("Sibling Differences in SES and Flu Vaccinations",
                 gp=gpar(fontsize=20,font=3)))

```

## Within Family

```{r plot-within-family, echo=TRUE, message=FALSE}
# Within Family
## Setup: Use discord_data


discord_flu <- discord_data(
  data = flu_modeling_data,
  outcome = "flu_total",
  predictors = "s00_h40",
  id = "extended_id",
  sex = "sex",
  race = "race",
  pair_identifiers = c("_s1", "_s2"),
  demographics = "both"
) %>%
  filter(!is.na(s00_h40_mean), !is.na(flu_total_mean)) %>%
  mutate(
    # # Classify Difference Grouping
    ses_diff_group = factor(case_when(s00_h40_diff > 0.33 ~ "More Advantage",
                                      s00_h40_diff < -0.33 ~ "Less Advantage",
                                      abs(s00_h40_diff) <= 0.33 ~ "Equally Advantage"),
                            levels = c("Less Advantage",
                                       "Equally Advantage",
                                       "More Advantage"))
  )

shading <- c("firebrick4", "firebrick1", "dodgerblue1", "dodgerblue4")


# Color scale settings
max_val <- max(abs(discord_flu$s00_h40_diff), na.rm = TRUE)
values <- seq(-max_val, max_val, length = length(shading))
# Main scatter plot
main_plot <- ggplot(discord_flu, aes(x = s00_h40_diff,
                                     y = flu_total_diff, colour = s00_h40_diff)) +
  geom_point(size = 0.8, alpha = 0.8, na.rm = TRUE,
             position = position_jitter(w = 0.2, h = 0.2)) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  scale_colour_gradientn(
    name = "Sibling\nDifferences\nin SES",
    colours = shading,
    na.value = "#AD78B6",
    values = scales::rescale(c(min(discord_flu$flu_total_diff, na.rm = TRUE),
                               mean(discord_flu$flu_total_diff, na.rm = TRUE),
                               max(discord_flu$flu_total_diff, na.rm = TRUE)))
  ) +
  theme_minimal() +
  theme( legend.position = "left")+
  labs(x = "Diff SES at Age 40", y = "Diff Flu Vaccinations (2006–2016)")

main_plot
```
